---
title: "About Recipes"
author: "Ryan Garnett"
date: "2023-03-29"
categories: [about]
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
# Load packages
library(gt)
library(readr)
library(fontawesome)
library(purrr)
library(stringr)
library(tidyr)
library(dplyr)
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
# Functions, variables, and data
pepper_on <- fontawesome::fa("pepper-hot", fill = "#A62014", height = "2em")
pepper_off <- fontawesome::fa("pepper-hot", fill = "#9FC2B9", height = "2em")

spicy0 <- paste0(pepper_off, pepper_off, pepper_off)
spicy1 <- paste0(pepper_on, pepper_off, pepper_off)
spicy2 <- paste0(pepper_on, pepper_on, pepper_off)
spicy3 <- paste0(pepper_on, pepper_on, pepper_on)

colour1 <- "#153331"
colour2 <- "#1E6E6E"
```

The structure and design of this blog is based on iterative learning, starting with the most basic and build by adding one new element concept. Learning highly technical subjects can be challenging. A method to over come this challenge is to make a 1-to-1 comparison based on a four part idea:

1.  concept (what)
2.  purpose (why)
3.  structure (how)
4.  example (how)

<br>

> *Linear learning styles with the sole aim of becoming fluent in a language may result in the inability to perform basic and tangible actions desired by those wanting to communicate within that language* <br> Johnny Harris

Thinking of adult learners and individuals learning on the job who are looking for help on a specific challenge, the blog has been structured to be small easily consumable chunks similar to that of a recipe card. The concept for a recipe card is that they are self contained, providing all the ingredients, preparation, and instructions required to create a meal. While a cookbook may consist of many recipes, there is no expectation to read, understand, and master all the recipes in order to prepare a meal. Following this as the central theme the blog, it has been designed as a number of data analytics recipes focusing on the R language.

</br>

The blog is built around a four course meal starting at the very beginning for individuals with no prior experience coding R, but looking to join the R data analytics community. In R there are many ways to undertake and tackle a problem, each with their own pros and cons. The examples, or recipes, in this blog are by no means the only, best, or most efficient means of performing a certain data analytics task. The approach taken is to build a base that can be easily built on using a similar style and coding syntax.

The sections, courses, are independent of each other, allowing a learner to begin according to their individual R data analytics journey, organized into four sections:

1.  Appetizers
2.  Breakfast
3.  Lunch
4.  Supper

</br>

<p style="font-size: 1.5em;"> Spicyness </p>

I love spicy food; tacos, burritos, chili, curry, hot sauce, almost anything, and I add hot sauce to almost everything. When beginning to eat spicy food you generally do not start with a Carolina Reaper, ghost pepper, or a habanero; most people may start with adding Frank's Hot Sauce, a little chili powder, or even some fresh jalapenos to a meal. Similar to the progression of spicy food a spicy index was used to help communicate the relative difficulty of the analytical concept or technical coding difficulty.

::: recipe-desc
::: spicy
`r pepper_off` `r pepper_off` `r pepper_off`
:::

<b>Not spicy</b>: Not technically difficult subject matter
:::

::: recipe-desc
::: spicy
`r pepper_on` `r pepper_off` `r pepper_off`
:::

<b>Mild spicy</b>: Minimal technically difficult subject matter
:::

::: recipe-desc
::: spicy
`r pepper_on` `r pepper_on` `r pepper_off`
:::

<b>Moderate spicy</b>: Moderate technically difficult subject matter
:::

::: recipe-desc
::: spicy
`r pepper_on` `r pepper_on` `r pepper_on`
:::

<b>Very spicy</b>: Technically difficult subject matter
:::
